---
title: 'R Exercises Part 2: Data Frames and Other Data Structures'
author: "Lauren Gardiner"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The original document for the part 2 of the R exercises can be found (here)[https://nuitrcs.github.io/intro_r_workshop/coreexercises/part2_exercises.html]

# Exercise: Make a `data.frame`

```{r}
mydata <- data.frame(var1=1:10, var2=LETTERS[1:10], var3=c(6, 3, 2, 7, 9, 11, 73, 12, 83, 29))
mydata
```
```{r}
mydata$month <- month.name[1:10]
mydata
```

# Exercise: Explore Differences in Selecting from `data.frame`
```{r}
mydata[1]
mydata[[1]]
mydata$var2
mydata["var2"]
mydata[1, 1]
mydata[, 1]
mydata[1,]
mydata[-1,]
```
# Exercise: Working with a `data.frame`
```{r}
data(iris)
dim(iris)
```
```{r}
names(iris)
```
```{r}
head(iris, 10)
```
```{r, eval = FALSE}
View(iris)
```
```{r}
iris[iris$Sepal.Width < 2.5,]
```
```{r}
iris[iris$Sepal.Width < 2.5 & iris$Sepal.Length < 5,]
```
```{r}
iris[which.max(iris$Sepal.Length),]
```
```{r}
names(iris)[names(iris)=="Petal.Width"] <- "petalwidth"
names(iris)
```


# Exercise: Make a `list`
```{r}
mysubsets <- list(shortpetals=iris[iris$Petal.Length < 3,],
                  shortsepals=iris[iris$Sepal.Length < 5,])
mysubsets
```
```{r}
table(mysubsets$shortsepals$Species)
```

# Exercise: Fix subsetting errors
```{r}
iris[iris$Species=="setosa",]
```
# Exercise: Counting with Conditionals
```{r}
data(mtcars)
sum(mtcars$mpg > 30)
```
```{r}
sum(mtcars$hp < 100)
```

# Challenge Exercise: Matrix Manipulation
```{r}
X <- matrix(rnorm(50), nrow=10)
y <- 1:5
t(X %*% y)
```
```{r}
X <- matrix(rnorm(36), nrow=6)
solve(X)
```
```{r}
diag(solve(X))
```
# Challenge Exercise: List Indexing
```{r}
nested_list <- list(level1 = list(level2 = list(letters = LETTERS)))
nested_list$level1$level2$letters
```
```{r}
nested_list[[1]][[1]][[1]]
```
# Challenge Exercise: Use `apply`
```{r}
apply(mtcars, 2, mean)
```
