---
title: "Bootcamp Day 3"
author: "Lauren Gardiner"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Data Visualization Workshop
## Gapminder Data
### Setup
Load libraries
```{r}
library(tidyverse)
library(gapminder)
```
Load data
```{r}
data(gapminder)
schooldata <- read_csv("~/team/bootcamp/R/schooldata.csv")
```
### Base R
Create histogram
```{r}
hist(schooldata$grade)
```
You can change the bins with the `breaks` paramter
```{r}
hist(schooldata$grade, breaks=30)
```
You can also set the scales, axis limits, and titles
```{r}
hist(schooldata$grade, 
     breaks=seq(.875, 4.125, .25), 
     xlim=c(0,4), 
     main="Grade Distribution",
     xlab="Grade",
     ylab="Count")
```
Create a scatterplot of life expectancy on the y-axis and GDP per capita on the x-axis
```{r}
plot(x=gapminder$gdpPercap[gapminder$year==2002],
     y=gapminder$lifeExp[gapminder$year==2002],
     log="x")
```
To save a plot, you need to start a device with a file name, recreate the graphics, and then save it. You can set dimensions, which are in inches. To save the graphic, turn off the use the `dev.off()` command
```{r}
pdf("plot1.pdf", width=6, height=4)
plot(x=gapminder$gdpPercap[gapminder$year==2002],
     y=gapminder$lifeExp[gapminder$year==2002],
     log="x")
dev.off()
```
### ggplot2
Let's recreate all of the base R plots in `ggplot`

Create a histogram with `ggplot`
```{r}
ggplot(schooldata, aes(grade)) + 
  geom_histogram(bins=20)
```
Instead of setting the number of bins with `bins`, you can set the `binwidth`
```{r}
ggplot(schooldata, aes(grade)) + 
  geom_histogram(binwidth=.25) +
  xlim(0, NA) +
  xlab("Grade") +
  ylab("Count") +
  ggtitle("Grade Distribution")
```
Let's split the data by gender. ggplot's `facet` handles this much better than base R where you have to set the layout and make each individual plot
```{r}
ggplot(schooldata, aes(grade)) + 
  geom_histogram(binwidth=.25) +
  xlim(0, NA) +
  xlab("Grade") +
  ylab("Count") +
  ggtitle("Grade Distribution") +
  facet_grid(gender~.)
```
Let's filter out the missing data and set the order with `dplyr` before handing it over to `ggplot`
```{r}
schooldata %>%
  filter(!is.na(gender)) %>%
  mutate(gender = factor(gender, 
                         levels=c("M", "F"),
                         labels=c("Male", "Female"))) %>%
  ggplot(aes(grade)) + 
  geom_histogram(binwidth=.25) +
  xlim(0, NA) +
  xlab("Grade") +
  ylab("Count") +
  ggtitle("Grade Distribution") +
  facet_grid(gender~.)
```
Create a scatterplot of life expectancy on the y-axis and GDP per capita on the x-axis and color by the variable `continent`
```{r}
gapminder %>%
  filter(year==2002) %>%
  ggplot(aes(x=gdpPercap, y=lifeExp, color=continent)) +
  geom_point() +
  scale_x_log10()
```
Save the plot. `ggsave` will save the last plot that was made. 
```{r}
ggsave("p1.pdf", width=6, height=4)
```
## Baby Names Data
### Setup
```{r}
library(babynames)
data(babynames)
```
### Plotting with `ggplot`
Create initial and initial proportions columns
```{r}
babynames <- mutate(babynames, initial=substr(name, 1, 1))
initial_props <- 
  babynames %>%
  group_by(year, sex, initial) %>%
  summarize(count=sum(n), totalprop=sum(prop)) %>%
  ungroup()
```
Plot girls with the initial `A` over time
```{r}
initial_props %>%
  filter(sex=="F" & initial=="A") %>%
  ggplot(aes(y=totalprop, x=year)) +
  geom_line()
```
Plot boys and girls with the initial `A` over time
```{r}
initial_props %>%
  filter(initial=="A") %>%
  ggplot(aes(y=totalprop, x=year, color=sex)) +
  geom_line()
```
Change the `linetype` instead of `color`
```{r}
initial_props %>%
  filter(initial=="A") %>%
  ggplot(aes(y=totalprop, x=year)) +
  geom_line(aes(linetype=sex), color="red", size=1)
```
Create plot of proportion by year for each initial
```{r}
initial_props %>%
  ggplot(aes(y=totalprop, x=year, color=sex)) +
  geom_line() +
  facet_wrap(~initial) +
  scale_color_discrete(labels=c("Female", "Male"), name="Sex") +
  scale_y_continuous(breaks=c(0, .1, .2)) +
  theme(axis.text.x = element_text(angle=-90, vjust=0.5),
        panel.grid.minor = element_blank()) +
  labs(title="Baby Names First Initial",
       x="Year",
       y="Proportion of Babies")
```
#### Let's find unisex names with `dplyr`
```{r}
topnames <- babynames %>%
  filter(year>=1950) %>%
  group_by(name) %>%
  summarize(total=sum(n)) %>%
  arrange(desc(total)) %>%
  slice(1:1000) %>%
  inner_join(babynames) %>%
  filter(year >= 1950) %>%
  mutate(sex=recode(sex, "F"="Female", "M"="Male")) %>%
  group_by(name, sex, total) %>%
  summarise(sextotal=sum(n)) %>%
  spread(key=sex, value=sextotal) %>%
  mutate(ratio=(Male-Female)/total)
```
Create a scatterplot with `topnames` dataframe
```{r}
ggplot(topnames, aes(x=Male, y=Female)) +
  geom_point()
```
Find row where there is missing data
```{r}
topnames %>%
  filter(is.na(Male) | is.na(Female))
```
Set limits to focus on bottom right part of the plot
```{r}
ggplot(topnames, aes(x=Male, y=Female, color=ratio)) +
  geom_point() +
  lims(x=c(0, 250000), y=c(0, 250000)) +
  geom_abline() +
  scale_color_gradient(low="pink", high="blue", 
                       name="Sex Dominance", 
                       breaks=c(.9, 0, -.9), 
                       labels=c("Male", "Neutral", "Female")) +
  geom_text(aes(label=ifelse(abs(ratio) < .2, 
                                 name,
                                 "")),
            hjust=-.25, vjust=0.5, color="gray10",
            fontface="bold", size=3)
```
Reconfigure the data to be used with a bar chart. For a bar chart, you need to gather the data and have groups. (This is undoing what we did with `spread` when creating `topnames`)
```{r}
topnames2 <- gather(topnames, key="sex", value="sextotal", Male, Female)
```
Create a bar chart for the most common unisex names
```{r}
topnames2 %>%
  filter(abs(ratio) < .3 & !is.na(ratio)) %>%
  ggplot(aes(x=name, y=sextotal, fill=sex)) +
  geom_bar(stat="identity") +
  labs(title="Popular Unisex Names", x="", y="Count")
```
Show the split instead of count
```{r}
topnames2 %>%
  filter(abs(ratio) < .3 & !is.na(ratio)) %>%
  ggplot(aes(x=name, y=sextotal, fill=sex)) +
  geom_bar(stat="identity", position="fill") +
  geom_hline(yintercept = .5) +
  labs(title="Popular Unisex Names", x="", y="Split")
```
Show a paired bar chart instead of stacked bar chart
```{r}
topnames2 %>%
  filter(abs(ratio) < .3 & !is.na(ratio)) %>%
  ggplot(aes(x=name, y=sextotal, fill=sex)) +
  geom_bar(stat="identity", position="dodge") +
  labs(title="Popular Unisex Names", x="", y="Count")
```
Load `ggthemes` package
```{r}
library(ggthemes)
```
Change theme of plot
```{r}
topnames2 %>%
  filter(abs(ratio) < .3 & !is.na(ratio)) %>%
  ggplot(aes(x=name, y=sextotal, fill=sex)) +
  geom_bar(stat="identity") +
  labs(title="Popular Unisex Names", x="", y="Count") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

```{r}
initial_props %>%
  filter(initial=="A") %>%
  ggplot(aes(y=totalprop, x=year, color=sex)) +
  geom_line() +
  theme_fivethirtyeight()
```
Add theme after plot and save
```{r}
myplot <- initial_props %>%
  filter(initial=="A") %>%
  ggplot(aes(y=totalprop, x=year, color=sex)) +
  geom_line()
myplot + theme_calc()

ggsave("plot2.pdf", myplot)
```
### Plotting with `plotly`
Load plotly library
```{r}
library(plotly)
```
Create column with first two letters of each name and subset `A` female names
```{r}
babynames <- mutate(babynames, first2=substr(name, 1, 2))
anames <- babynames %>%
  filter(initial=="A" & sex=="F") %>%
  group_by(year, first2) %>%
  summarize(count=sum(n), totalprop=sum(prop)) %>%
  ungroup()
```
Plot `anames` proportion by year
```{r}
p1 <- anames %>%
  ggplot(aes(y=totalprop, x=year, color=first2)) +
  geom_line()
```
Use wrapper function to be able to get tooltips with `plotly`
```{r}
ggplotly(p1)
```
Make a plot directly with `plotly`
```{r}
plot_ly(anames, x=~year, y=~totalprop, color=~first2, type="scatter", mode="lines")
```
Make a plot 3D (poor example due to data, but shows functionality)
```{r}
plot_ly(anames, x=~year, y=~first2, z=~totalprop, color=~first2, line=list(width=4)) %>%
  add_lines() %>%
  layout(scene=list(xaxis=list(title="Year"),
                    yaxis=list(title="Starting Letters"),
                    zaxis=list(title="Proportion")))
```