---
title: "Stats Example"
author: "Lauren Gardiner"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

Get data

```{r}
data(mtcars)
head(mtcars)
```

# Basic Stats

Calculate summary statistics

```{r}
mean(mtcars$mpg)
summary(mtcars$mpg)
```

Create a simple table

```{r}
table(mtcars$gear)
```

Create a table with two variables

```{r}
table(mtcars$gear, mtcars$am)
```
Create a table with more than two variables
```{r}
ftable(mtcars$gear, mtcars$am, mtcars$cyl)
```

Create a proportion table

```{r}
prop.table(table(mtcars$gear))
```

```{r, eval = FALSE}
mtcars %>%
  group_by(gear) %>%
  summarize(mean(mpg),
            mean(hp))
```

# Correlation

Calculate correlation between two variables

```{r}
cor(mtcars$mpg, mtcars$disp)
```

Calculate pairwise correlation for an entire dataframe

```{r}
cor(mtcars)
```

Calculate covariance

```{r}
cov(mtcars)
```

# t-test

Do automatic transmission cars have different mpg than manual transmission cars?

```{r}
t.test(mtcars$mpg[mtcars$am == 0], mtcars$mpg[mtcars$am == 1])
```

You can also use a formula format instead of passing two vectors of values

```{r}
t1 <- t.test(mtcars$mpg ~ mtcars$am)
names(t1)
```

You can pull out a specific value from the `t.test`

```{r}
t1$p.value
```

# Linear Model

The typical format for a linear model is as follows:

```{r, eval=FALSE}
lm(y ~ x1 + x2 + x3, data=df)
```

Here is an example with our `mtcars` data:

```{r}
r1 <- lm(mpg ~ hp, data = mtcars)
summary(r1)
```

Model results and summary have different info

```{r}
names(r1)
```

```{r}
names(summary(r1))
```

We can plot the model and use `identify` to label specific points (need to use `identify` in console or script instead of markdown)

```{r}
plot(mtcars$mpg ~ mtcars$hp)
abline(r1)
identify(mtcars$hp, mtcars$mpg, labels = rownames(mtcars))
```

The following command will show you four plots in the console

```{r, eval=FALSE}
plot(r1)
```

Create a second linear model

```{r}
r2 <- lm(mpg ~ hp + I(hp^2) + am + cyl, data=mtcars)
summary(r2)
```

`poly` will give you each power up until the one specified, in this case 2. 

```{r}
r2 <- lm(mpg ~ poly(hp, 2) + am + cyl, data=mtcars)
summary(r2)
```

`-1` will remove the intercept. `I(factor(variable)) is used for categorical data

```{r}
r3 <- lm(mpg ~ -1 + poly(hp, 2) + am + I(factor(cyl)), data=mtcars)
summary(r3)
```

`*` allows you to have interaction terms

```{r}
r4 <- lm(mpg ~ hp*am, data=mtcars)
summary(r4)
```

Create a `termplot`

```{r}
termplot(r2, rug=TRUE)
```

Predict mpg values

```{r}
r4 <- lm(mpg ~ -1 + poly(hp, 2) + am + I(factor(cyl)), data=mtcars)
newvals <- data.frame(hp=seq(50, 250, 10),
                      am=0,
                      cyl=4)
mpg_predict <- predict(r4, newvals)
mpg_predict
```

Visualize predicted values

```{r}
plot(mpg_predict ~ newvals$hp,
     type="l", lwd=3, col="red",
     main="am=0; cyl=4") 
points(mtcars$mpg ~ mtcars$hp)
```

